---
title: "Assignment Part 1"
author: "Kaumadi Wijesooriya (219362267)"
theme: cosmo
output:
  word_document: default
  pdf_document: default
  html_document: default
---
#github url: https://github.com/Kaumadi55/SLE-712.git
    
## Question 1- downloading file and reading into R
Here 'download.file()' command was used to obtain the dataset. 
  it was read into R. It is important to set 'header=TRUE' and 'raw.names=1', so that the data is formated properly. 
  Also the 'head' command was used to obtain first 6 lines of the dataset and 'str' command to obtain the structure. 
  The argument 'stringsAsFactors=FALSE' was used to obtain the factors as neumarical. 

```{r,chunk1,echo=TRUE}

# question 1

download.file("https://raw.githubusercontent.com/markziemann/SLE712_files/master/bioinfo_asst3_part1_files/gene_expression.tsv",destfile = "gene_expression.tsv")
readit<- read.table("gene_expression.tsv")
head(readit)
str(readit)
readit<- read.table("gene_expression.tsv",header = TRUE)
head(readit)
str(readit)
readit<- read.table("gene_expression.tsv",header = TRUE, stringsAsFactors = FALSE)
head(readit)
str(readit)
readit<- read.table("gene_expression.tsv",header = TRUE, stringsAsFactors = FALSE, row.names = 1)
head(readit)
str(readit)

```
## Question 2- Finding the mean
Here a new column has made for mean values of 2 RNA sequences, using the command 'mean()'. also '$' sign was used to perform the extraction as an object. Here the 'rowMeans()' command was used to obtain the mean values in a new row.

```{r,chunk2,echo=TRUE}

# question 2

readit$mean <- (readit$SRR5150592+readit$SRR5150593)/2
rowMeans(readit)
head(readit)

```
## Question 3- making an order for 10 genes with the highest mean expression
Here the 'order()' command was used to make the mean expressions in accending order and the minus sign indicates descending order.

```{r,chunk3,echo=TRUE}
# question 3

order(readit$mean)
order(-readit$mean)
readit[order(-readit$mean),]
expression <- readit[order(-readit$mean),]
head(expression,10)

```
## Question 4 - Determining the number of genes with a mean values <10
Here the 'subset()' command was used to make a subset of mean values less than 10. 
'nrow()' command was used to display the number of rows present in the data matrix.

```{r,chunk4,echo=TRUE}
# question 4

filtered<- subset(readit, mean < 10)
nrow(filtered)

```
## Question 5 - Making a histogram plot of the mean values
Here the 'hist()' command was used to compute a histogram for mean values. In the histogram, 'main' stands for the title of histogram. 
'xlab=' command represents the cordinates of 'X' axis and 'xlim=' command reprecents the cordinates of 'Y' axis. 
The 'border=' function was used to colour the border in black and 'col=' function was used to colour the column in green in the histogram. 
Also the 'breaks=' command use to obtain 30  break points in histogram.

```{r,chunk5,echo=TRUE}
# question 5

hist(readit$mean, main = "Histogram for mean values", xlab = "mean value", border = "black", col = "green", breaks = 30, xlim = c(0,50000))

```

## Question 6 - Downloading the tree growth data file and read it into R.
Here the 'download.file()' command was used to obtain the data file. Then it was read into R. 
It is important to set 'header=TRUE' so that the data is formated properly. 
Also the 'head' command was used to obtain first 6 lines of the dataset and 'str' command to obtain the structure.
the argument 'stringsAsFactors=FALSE' was used to obtain the factors as neumarical. 
Also 'colnames()' command was used to indicate column identification.

```{r,chunk6,echo=TRUE}
# question 6

download.file("https://raw.githubusercontent.com/markziemann/SLE712_files/master/bioinfo_asst3_part1_files/growth_data.csv", destfile = "growth_data.csv")
treegrowth <- read.table("growth_data.csv", sep = ",", header = TRUE)
treegrowth <- read.table("growth_data.csv", sep = ",", header = TRUE, stringsAsFactors = FALSE)
head(treegrowth)
str(treegrowth)
colnames(treegrowth)

```
## Question 7 - Calculating the mean and standard deviation of tree circumference
Here 2 subsets of data related to the tree growth from two sites (northeast and southwest) were obtained using 'subset()' command. 
Also 'mean()' and 'sd()' commands were used to obtain means and standard deviations of tree circumference at the begining and the end of tree growth.

```{r,chunk7,echo=TRUE}
# question 7

ne<- subset(treegrowth, Site =="northeast")
head(ne)
tail(ne)
mean(ne$Circumf_2004_cm)
mean(ne$Circumf_2019_cm)
sd(ne$Circumf_2004_cm)
sd(ne$Circumf_2019_cm)

sw<- subset(treegrowth, Site =="southwest")
head(sw)
tail(sw)
mean(sw$Circumf_2004_cm)
mean(sw$Circumf_2019_cm)
sd(sw$Circumf_2004_cm)
sd(sw$Circumf_2019_cm)

```
## Question 8 - Making a box plot of tree circumference
Here the 'boxplot()' function was used in order to make a box plot using tree circumference results of two cites at the begining and the end of growth.

```{r,chunk8,echo=TRUE}
# question 8

boxplot(ne$Circumf_2004_cm,ne$Circumf_2019_cm,sw$Circumf_2004_cm,sw$Circumf_2019_cm, 
        names = c("ne2004","ne2019","sw2004","sw2019"),ylab="Circumference(cm)", 
        main="Tree circumference at the start and end of the study")

```


## Question 9 - Calculating the mean growth
Here the 'mean()' command was used to calculate the mean growth of each site over the past 10 years.

```{r,chunk9,echo=TRUE}
# question 9

ne$treegrowth<- ne$Circumf_2019_cm-ne$Circumf_2009_cm
head(ne)
mean(ne$treegrowth)
sw$treegrowth<- sw$Circumf_2019_cm-sw$Circumf_2009_cm
head(sw)
mean(sw$treegrowth)

```
## Question 10 - Estimating the p value using t-test and wilcox test
Here, 't.test()' function and 'wilcox.test()' function were used to obtain p values in order to check wether there is a difference between tree circumference in the begining and the end of tree growth.

```{r,chunk10,echo=TRUE}
# question 10

t.test(sw$treegrowth,ne$treegrowth)
wilcox.test(sw$treegrowth,ne$treegrowth)
